Graph Isomorphism is Polynomially Equivalent to Counting Isomorphisms

### 2016-07-13 tex/below.tex
13 jul

We define $\totp$ using the following function. For each non-deterministic Turing machine $M$ and $x\in\{0,1\}^*$,
\[
	\tot_M(x) = (\text{\#paths of $M$ on input $x$}) - 1.
\]
A function $f:\{0,1\}^*\to\N$ is in $\totp$ if there exists a polynomial-time non-deterministic Turing machine $M$ such that $f = \tot_M$.

We define $\spanl$ using the following function. For each non-deterministic Turing machine $M$ and $x\in\{0,1\}^*$,
\[
	\sp_M(x) = \vert\{w \mid \text{$w$ is an output of $M$ on input $w$}\}\vert.
\]
A function $f:\{0,1\}^*\to\N$ is in $\spanl$ if there exists a log-space non-deterministic Turing machine $M$ such that $f = \sp_M$.

Given a language $L$, $\totp^{L}$ is defined by all functions $f:\{0,1\}^*\to\N$ such that $f = tot_{M^L}$.
\\

By the following result, we note that $\shp$ can be considered an extension of $\totp$.

\begin{theo}
	$\spanl \subseteq \totp$.
\end{theo}
\begin{proof}
	Let $f:\{0,1\}^*\to\N$ be a function in $\spanl$ and let $M$ be a non-deterministic log-space Turing machine such that $f = \sp_M$. We will provide a polynomial-time non-deterministic Turing machine $M'$ such that for each $x\in\{0,1\}^*$ it holds that $\sp_M(x) = \tot_{M'}(x)$. First we note that given a configuration $s$ of $M$ on input $x$ checking whether there exists a final state of $M$ which is reachable from $s$ can be done in polynomial time. This implies that checking whether $f(x) > 0$ can also be done in polynomial time. We present this machine $M'$ as a non-deterministic algorithm.

	\begin{algorithm}	
	\caption{Compute $f(x)+1$ paths}
	\begin{algorithmic}[]
		\Begin
		\If{$f(x) = 0$} stop
		\Else \text{ choose from}
		\State stop
		\State call $\textsc{GenTree}_{f}(x,\{s_0\},\epsilon)$
		\EndIf
		\EndBegin
		\Procedure{$\textsc{GenTree}_{f}(x,S,y)$}{}
		\State $S_0 = \emptyset$, $S_1 = \emptyset$
		\For{$s\in S$}
		\State do a search in the configuration graph of $M$ with input $x$ from $s$:
		\If{$s'$ is reached after writing 0 in the output tape} $S_0 \gets s'$
		\EndIf
		\If{$s'$ is reached after writing 1 in the output tape} $S_1 \gets s'$
		\EndIf
		\EndFor
		\If{a final state from $M$ is reachable from some $s\in S_0$ and some $s'\in S_1$} choose between:
		\State call $\textsc{GenTree}_{f}(x,S_0,y0)$
		\State call $\textsc{GenTree}_{f}(x,S_1,y1)$
		\ElsIf{a final state from $M$ is reachable from some $s\in S_i$ but not from any $s'\in S_{1-i}$}
		\State call $\textsc{GenTree}_{f}(x,S_i,yi)$
		\ElsIf{no final state from $M$ is reachable from any $s\in S_0 \cup S_1$}
		\Return $y$
		\EndIf
		\EndProcedure
	\end{algorithmic}
	\end{algorithm}

	For each output $y$ consider some computation path $p$ of $M$ with input $x$ that outputs $y$. Note there is at least one computation path in $M'$ that follows $p$ and outputs $y$. Also, note that each distinct computation path of $M'$ outputs a different $y$, since in each branch the word differs in the last bit. Taking into account the dummy path in line 4, we conclude that the number of computation paths of $M'$ with input $x$ is equal to $\sp_M(x)+1$.

\end{proof}

Given a deterministic machine $M$ we define the function $f_M\{0,1\}^*\to\N$ as:
\[
	f_M(x) = \text{the output of $M$ on input $w$}
\]
for each $w\in\{0,1\}^*$. A function $f:\{0,1\}^*\to\N$ is in FP if there exists a deterministic polynomial-time machine $M$ such that $f = f_M$.
\begin{theo} \label{diff}
	For each function $f\in\shp$ there exists functions $g\in \totp$ and $h\in\text{FP}$ such that $f = g - h$.
\end{theo}
\begin{proof}
	Let $f\in\shp$, and let $r:\{0,1\}^*\to\{0,1\}^*$ be such that for each $w\in\{0,1\}^*$ it holds that $f(w) = \shsat(r(w))$, and be computed in polynomial time. Let $\varphi_w = r(w)$.

@@

	We will now show that $g(w) = \tot_{M'}(w)$. Let $w\in\{0,1\}$. If $\tot_{M}(w) > p(\vert w \vert)$, then clearly $\tot_{M'}(w) = \tot_{M}(w) - p(\vert w \vert)$. If $\tot_{M}(w) \leq p(\vert w \vert)$, then it only computes the first simulation path and $\tot_{M'}(w) = 0$.
\end{proof}

@@

We present a counting class diagram:

@@

Or new addition to the known diagram is the following inclusion.

\begin{theo}
	$\spanp \subseteq \totp^{\np}$
\end{theo}
\begin{proof}
	Let $f\in\spanp$ and let $M$ be the non-deterministic polynomial-time machine for which $f = \sp_M$. We suppose without loss of generality that each output of $M(x)$ has length $p(\vert x \vert)$, where $p$ is a polynomial. We will provide a non-deterministic machine $M'^L$ where $L$ is the following language:
	\begin{multline*}
	L = \{(M,x,w_1,w_2) \mid \text{$M$ is a non-deterministic polynomial-time machine and} \\ \text{$M(x)$ gives some output in the lexicographic range $[w_1,w_2]$.} \}
	\end{multline*}
	The machine $M$ on input $x$ writes $w_1 = 0^{p(\vert x \vert)}, w_2 = 01^{p(\vert x \vert)-1}$, and asks the oracle $(M,x,w_1,w_2)$. If it answers yes, it generates a new branch and proceeds, if not, just proceeds. Then, write $w_1 = 10^{p(\vert x \vert)-1}, w_2 = 1^{p(\vert x \vert)}$ and asks the oracle $(M,x,w_1,w_2)$. If it answers yes, it generates a new branch and proceeds, if not, stop. TO BE CONTINUED.
\end{proof}

@@

>>>>

We define $\totp$ using the following function. For each non-deterministic Turing machine $M$ and $x\in\{0,1\}^*$,
$$
	\tot_M(x) = (\text{\#paths of $M$ on input $x$}) - 1.
$$
A function $f:\{0,1\}^*\to\N$ is in $\totp$ if there exists a polynomial-time non-deterministic Turing machine $M$ such that $f = \tot_M$.

For a non-deterministic Turing machine $M$ with an output tape, we define $\out_M(x)$ as the set of the outputs of each non-deterministic computation of $M$ on input $x$.

We define $\spanl$ using the following function. For each non-deterministic Turing machine $M$ and $x\in\{0,1\}^*$,
\[
	\sp_M(x) = \vert\out_M(x)\vert.
\]
A function $f:\{0,1\}^*\to\N$ is in $\spanl$ if there exists a log-space non-deterministic Turing machine $M$ such that $f = \sp_M$.

Similarly, a function  $f:\{0,1\}^*\to\N$ is in $\spanp$ if there exists a polynomial-time non-deterministic Turing machine $M$ such that $f = \sp_M$.

Given a language $L$, $\totp^{L}$ is defined by all functions $f:\{0,1\}^*\to\N$ such that $M$ is a non-deterministic polynomial machine and $f = \tot_{M^L}$.
\\

Now we will show some properties of $\totp$.

Given a deterministic machine $M$ we define the function $f_M\{0,1\}^*\to\N$ as:
\[
	f_M(x) = \text{the output of $M$ on input $w$}
\]
for each $w\in\{0,1\}^*$. A function $f:\{0,1\}^*\to\N$ is in FP if there exists a deterministic polynomial-time machine $M$ such that $f = f_M$.

The following proof is added as a completion of a theorem of Pagourtzis and Zachos \cite{DBLP:conf/mfcs/PagourtzisZ06}.
\begin{theo} \label{diff}
	For each function $f\in\shp$ there exist functions $g\in \totp$ and $h\in\text{FP}$ such that $f = g - h$.
\end{theo}
\begin{proof}
	Let $f\in\shp$, and let $r:\{0,1\}^*\to\{0,1\}^*$ be such that for each $w\in\{0,1\}^*$ it holds that $f(w) = \shsat(r(w))$, and be computed in polynomial time. Let $\varphi_w = r(w)$.

@@

	We will now show that $g(w) = \tot_{M'}(w)$. Let $w\in\{0,1\}^*$. If $\tot_{M}(w) > p(\vert w \vert)$, then clearly $\tot_{M'}(w) = \tot_{M}(w) - p(\vert w \vert)$. If $\tot_{M}(w) \leq p(\vert w \vert)$, then it only computes the first simulation path and $\tot_{M'}(w) = 0$.
\end{proof}

@@

We also define the class $\maxp$ (resp. $\minp$) as the class of functions $f:\{0,1\}^* \to \N$ where there exists some non-deterministic polynomial-time machine $M$ such that for each $x\in\{0,1\}^*$ it holds that $f(x) = \max(\out_M(x))$ (resp. $\min(\out_M(x))$).

We present an updated class diagram of the classes surrounding $\totp$.

@@

Inclusions are indicated with arrows. Separation results under complexity assumptions are indicated with dashed arrows.

The inclusion $\shl \subseteq \spanl$ and the fact that $\spanl \subseteq \fp$ implies $\ph = \text{P}^{\np}$ were showed by Àlvarez and Jenner $\cite{DBLP:journals/tcs/AlvarezJ93}$. The inclusions $\fp \subseteq \totp$ and $\totp \subseteq \shp$, and the fact that $\shp \subseteq \totp$ implies $\text{P} = \np$ were showed by Pagourtzis and Zachos in \cite{DBLP:conf/mfcs/PagourtzisZ06}. The inclusions $\fp^{\np} \subseteq \totp^{\np}$ and $\totp^{\np} \subseteq \shnp = \shp^{\np}$ are trivial relativizations. The inclusions $\minp \subseteq \fp^{\np}$, $\maxp \subseteq \fp^{\np}$, $\maxp \subseteq \spanp$ and $\shp \subseteq \spanp$, and the fact that $\spanp \subseteq \shp$ implies that $\np = \up$ were showed by Kobler et. al. in \cite{DBLP:journals/acta/KoblerST89}. The inclusion $\fp \subseteq fp^{\np}$ is trivial. The inclusion $\shl \subseteq \fp$ was showed in [COMPLETE THIS].

\begin{theo}
	$\spanl \subseteq \totp$.
\end{theo}
\begin{proof}
	Let $f:\{0,1\}^*\to\N$ be a function in $\spanl$ and let $M$ be a non-deterministic log-space Turing machine such that $f = \sp_M$. We will provide a polynomial-time non-deterministic Turing machine $M'$ such that for each $x\in\{0,1\}^*$ it holds that $\sp_M(x) = \tot_{M'}(x)$. First we note that given a configuration $s$ of $M$ on input $x$ checking whether there exists a final state of $M$ which is reachable from $s$ can be done in polynomial time. This implies that checking whether $f(x) > 0$ can also be done in polynomial time. We present this machine $M'$ as a non-deterministic procedure in Algorithm \ref{spanltot}.

	\begin{algorithm}	
		\caption{Compute $f(x)+1$ paths} \label{spanltot}
		\begin{algorithmic}[]
			\Begin
			\If{$f(x) = 0$} stop
			\Else \text{ choose from}
			\State stop
			\State call $\textsc{GenTree}_{f}(x,\{s_0\},\epsilon)$
			\EndIf
			\EndBegin
			\Procedure{$\textsc{GenTree}_{f}(x,S,y)$}{}
			\State $S_0 = \emptyset$, $S_1 = \emptyset$
			\For{$s\in S$}
			\State do a search in the configuration graph of $M$ with input $x$ from $s$:
			\If{$s'$ is reached after writing 0 in the output tape} $S_0 \gets s'$
			\EndIf
			\If{$s'$ is reached after writing 1 in the output tape} $S_1 \gets s'$
			\EndIf
			\EndFor
			\If{a final state from $M$ is reachable from some $s\in S_0$ and some $s'\in S_1$} choose between:
			\State call $\textsc{GenTree}_{f}(x,S_0,y0)$
			\State call $\textsc{GenTree}_{f}(x,S_1,y1)$
			\ElsIf{a final state from $M$ is reachable from some $s\in S_i$ but not from any $s'\in S_{1-i}$}
			\State call $\textsc{GenTree}_{f}(x,S_i,yi)$
			\ElsIf{no final state from $M$ is reachable from any $s\in S_0 \cup S_1$}
			\Return $y$
			\EndIf
			\EndProcedure
		\end{algorithmic}
	\end{algorithm}

	For each output $y \in \out_M(x)$ consider some computation path $p$ of $M$ with input $x$ that outputs $y$. Note there is at least one computation path in $M'$ that follows $p$ and outputs $y$. Also, note that each distinct computation path of $M'$ outputs a different $y$, since in each branch the word differs in the last bit. Taking into account the dummy path in line 4, we conclude that the number of computation paths of $M'$ with input $x$ is equal to $\sp_M(x)+1$. 
\end{proof}

\begin{theo}
	$\spanp \subseteq \totp^{\np}$
\end{theo}
\begin{proof}
	Let $f\in\spanp$ and let $M$ be the non-deterministic polynomial-time machine for which $f = \sp_M$. We suppose without loss of generality that each output of $M(x)$ has length $p(\vert x \vert)$, where $p$ is a polynomial. We will provide a non-deterministic machine $M'^L$ where $L$ is the following language:
	\begin{multline*}
	L = \{(M,x,w_1,w_2) \mid \text{$M$ is a non-deterministic polynomial-time machine and} \\ \text{$M(x)$ gives some output in the lexicographic range $[w_1,w_2]$.} \}
	\end{multline*}
	This is presented as a non-deterministic procedure in Algorithm \ref{spanptotnp}.

		\begin{algorithm}	
			\caption{Compute $f(x)+1$ paths} \label{spanptotnp}
			\begin{algorithmic}[]
				\Begin
				\If{$f(x) = 0$} stop
				\Else \text{ choose from}
				\State stop
				\State call $\textsc{GenTree}_{f}(x,\varepsilon,0)$
				\EndIf
				\EndBegin
				\Procedure{$\textsc{GenTree}_{f}(x,w,m)$}{}
				\If{$m = p(\vert x \vert)$}
				\State stop
				\Else
				\State $w_1 = w0^{p(\vert x \vert)-m}$
				\State $w_2 = w01^{p(\vert x \vert)-m-1}$
				\If{$(M,x,w_1,w_2)\in L$}
				\State call $\textsc{GenTree}_{f}(x,w0,m+1)$
				\EndIf
				\State $w_1 = w10^{p(\vert x \vert)-m-1}$
				\State $w_2 = w1^{p(\vert x \vert)-m}$
				\If{$(M,x,w_1,w_2)\in L$}
				\State call $\textsc{GenTree}_{f}(x,w1,m+1)$
				\EndIf
				\EndIf
				\EndProcedure
			\end{algorithmic}
		\end{algorithm}

		Note that on input $x$, for each $w \in \out_M(x)$ there exists some branch in the algorithm that calls $\textsc{GenTree}_{f}(x,w,p(\vert x \vert))$ and then stops. Also note that for each branch in the algorithm that is not the dummy path in line 4, there is some $w$ in the last call to $\textsc{GenTree}_{f}$ that belongs to $\out_M(x)$. From this we conclude that $f(x) = \tot_{M'^L}(x)+1$.
\end{proof}

@@

We provide a characterization of $\shnp$ that doesn't use oracles and lets each function in $\shnp$ be defined by a single non-deterministic machine. We contrast this result with the one showed by Kobler et. al. in \cite{DBLP:journals/acta/KoblerST89} that lets define each function in $\shnp$ without oracles using two machines.
